\section{Proof Outline}

\begin{lemma}[Measurable space of expressions]
  Well-typed expressions of each type form a measurable space $(Expr_\tau, \mathcal{F}_{Expr_{\tau}})$.
\end{lemma}

\subsection{Small-step}

\begin{lemma}[Small-step semantics]
  The small-step semantics $\sem{e : \tau} \in \mathcal{D}(Expr_{\tau})$ is well-defined for all well-typed $e : \tau$.
\end{lemma}

\begin{lemma}[Type preservation]
  Let $e$ be a well-typed expression of type $\tau$, and assume that
  $\{e' \mid \emptyset \vdash e' : \tau'\}$ is measurable for every type $\tau'$.
  Then
  \[
    \sem{e : \tau}\bigl(\{\,e' \mid \emptyset \nvdash e' : \tau\,\}\bigr) = 0.
  \]
  That is, the small-step semantics assigns measure zero to the set of
  expressions that are not well-typed at type $\tau$.
\end{lemma}

\begin{lemma}[Step is a Markov kernel]
  The map $e \mapsto \sem{e : \tau}$ defines a Markov kernel $Expr_\tau \to \mathcal{D}(Expr_\tau)$.
\end{lemma}

\subsection{Big-step}

\begin{lemma}[$n$-step semantics]
  The $n$-step semantics $\bigsemstep{e : \tau}{n}$ is defined by iterating the small-step kernel $n$ times, and is itself a Markov kernel.
\end{lemma}

\begin{lemma}[Monotone accumulation of mass]
  For all $n \in \mathbb{N}$:
  \begin{enumerate}
    \item $\bigsemstep{e : \tau}{n}(Val_\tau) \leq \bigsemstep{e : \tau}{n+1}(Val_\tau)$
  \end{enumerate}
  That is, probability mass on values accumulates monotonically.
\end{lemma}

\begin{lemma}[Big-step semantics exists]
  The pointwise limit
  \[
    \bigsem{e : \tau} := \lim_{n \to \infty} \bigsemstep{e : \tau}{n}
  \]
  exists as a sub-probability measure on $Val_\tau$, by the monotone convergence theorem.
\end{lemma}

\subsection{Transformation}

\begin{definition}[Expectation transformation]
  Define $\exptrans{e : \mathsf{float}}$ inductively on the structure of $e$.

  *** This will be dependent on the types.
\end{definition}


\subsection{Expectations}

\begin{lemma}[Expectation is well-defined]
  For any sub-probability measure $\mu \in \mathcal{D}(Vals_{\tau})$, the expectation
  \[
    \mathbb{E}[\mu] := \int x \, d\mu(x)
  \]
  is a well-defined real number.

  **Specifically, let us take $\tau = \mathbf{float}$.
\end{lemma}

\begin{lemma}[Expectation of a Dirac]
  For any $c \in Vals_{\tau}$:
  \[
    \mathbb{E}[\delta_c] = c
  \]
\end{lemma}

\begin{lemma}[Linearity of expectation]
  For sub-probability measures $\mu_1, \mu_2 \in \mathcal{D}(Vals_{\tau})$ and $a, b \in \mathbb{R}$:
  \[
    \mathbb{E}[a \cdot \mu_1 + b \cdot \mu_2] = a \cdot \mathbb{E}[\mu_1] + b \cdot \mathbb{E}[\mu_2]
  \]
\end{lemma}

\begin{lemma}
  For $\mu \in \mathcal{D}(Expr_\tau)$ and measurable $K : Expr_\tau \to \mathcal{D}(Vals_{\tau})$:
  \[
    \mathbb{E}[\mu \monbind K] = \int \mathbb{E}[K(x)] \, d\mu(x)
  \]
\end{lemma}

\begin{lemma}[Expectation of a program is well-defined]
  For any well-typed expression $e : \mathsf{float}$, the expectation
  \[
    \mathbb{E}[\bigsem{e : \mathsf{float}}] 
  \]
  is a well-defined real number.
\end{lemma}

\subsection{Affine condition}

\begin{definition}[Affine continuation]
  A measurable function $K : \mathbb{R} \to \mathcal{D}(\mathbb{R})$ is called \emph{expectation-affine} if the map
  \[
    x \mapsto \mathbb{E}[K(x)]
  \]
  is an affine function of $x$, i.e.\ there exist $a, b \in \mathbb{R}$ such that $\mathbb{E}[K(x)] = ax + b$ for all $x$.
\end{definition}

\begin{itemize}
  \item The continuation $K(x) = \delta_{ax+b}$ is expectation-affine. 
  \item If $K_1 : \mathbb{R} \to \mathcal{D}(\mathbb{R})$ is expectation-affine and $K_2 : \mathbb{R} \to \mathcal{D}(\mathbb{R})$ is expectation-affine, then their monadic composition
  \[
    x \mapsto K_1(x) \monbind K_2
  \]
  is expectation-affine.
  \item The continuation $K(x) = \mathcal{N}(ax + b, \sigma^2)$ is expectation-affine, with $\mathbb{E}[K(x)] = ax + b$.
  \item The continuation $K(x) = \mathrm{Uniform}(ax + b,\, cx + d)$ is expectation-affine, with
  \[
    \mathbb{E}[K(x)] = \frac{(a+c)}{2} x + \frac{(b+d)}{2}.
  \]
\end{itemize}

\begin{lemma}
  For any $\mu \in \mathcal{D}(Expr_\mathsf{float})$ and any expectation-affine continuation $K$:
  \[
    \mathbb{E}[\mu \monbind K] = \mathbb{E}[\delta_{\mathbb{E}[\mu]} \monbind K]
  \]
\end{lemma}

\begin{proof}
  \begin{align*}
    &\mathbb{E}[\mu \monbind K] \\
    = \quad&\int \mathbb{E}[K(x)] \, d\mu(x) \tag{Lemma 5.12} \\
    = \quad&\int (ax + b) \, d\mu(x) \tag{Definition 5.14} \\
    = \quad&a \cdot \mathbb{E}[\mu] + b \tag{Linearity of expectation} \\
    = \quad&\mathbb{E}[K(\mathbb{E}[\mu])] \tag{Definition 5.14} \\
    = \quad&\mathbb{E}[\delta_{\mathbb{E}[\mu]} \monbind K] 
  \end{align*}
\end{proof}

\subsection{Soundness?}

\begin{theorem}[Soundness of expectation transformation]
  Let $e : \mathsf{float}$ be a well-typed expression. Then:
  \[
    \mathbb{E}[\bigsem{e : \mathsf{float}}] = \mathbb{E}[\bigsem{\mathcal{E}( e : \mathsf{float}) : \mathsf{float}}].
  \]

  We actually don't necessarily need all continuations arising in $e$ be expectation affine.

  And even if $\bigsem{e : \mathsf{float}}$ and $\bigsem{\mathcal{E}( e : \mathsf{float})}$ are different distributions, they can still have the same expectation.
\end{theorem}
